---
title: "Data Profiling Report"
output:
  html_document: 
    theme: readable
    toc: yes
params:
  data: data
---

```{r global_options, include=FALSE}
data <- params$data
knitr::opts_chunk$set(fig.width=14, fig.height=10, echo=FALSE)
```

### Basic Statistics
The data is **`r format(object.size(data), units="auto")`** in size. There are **`r format(nrow(data), big.mark=",")`** observations (rows) and **`r format(ncol(data), big.mark=",")`** features (columns), with **`r SplitColType(data)$num_discrete`** discrete and **`r SplitColType(data)$num_continuous`** continuous. There are **`r format(sum(is.na(data)), big.mark=",")`** missing values out of **`r format(nrow(data)*ncol(data), big.mark=",")`** data points.

Below is the structure of the data.
```{r}
str(data)
```

### Missing Values
The following graph shows the distribution of missing values.
```{r}
PlotMissing(data)
```

### Data Distribution
#### Discrete Features
The following graph shows the frequency distribution of all discrete features.
```{r }
BarDiscrete(data)
```

#### Continuous Features
###### Boxplot
The following graph shows the general distribution of all continuous features.
```{r}
BoxplotContinuous(data)
```

###### Histogram
The following graph shows the density distribution of all continuous features.
```{r}
HistogramContinuous(data)
```

