---
title: "Data Profiling Report"
output:
  html_document: 
    theme: cerulean
    toc: yes
    toc_depth: 6
params:
  data: data
  fun_options: fun_options
---

```{r global_options, include=FALSE}
data <- params$data
opts <- params$fun_options
knitr::opts_chunk$set(fig.width=14, fig.height=10, echo=FALSE)
```

### Basic Statistics
The data is **`r format(object.size(data), units = "auto")`** in size. There are **`r format(nrow(data), big.mark = ",")`** observations (rows) and **`r format(ncol(data), big.mark = ",")`** features (columns), with **`r SplitColType(data)$num_discrete`** discrete and **`r SplitColType(data)$num_continuous`** continuous. There are **`r format(sum(is.na(data)), big.mark = ",")`** missing values out of **`r format(nrow(data)*ncol(data), big.mark = ",")`** data points.

Below is the structure of the data.
```{r data_structure}
str(data)
```

### Missing Values
The following graph shows the distribution of missing values.
```{r plot_missing}
PlotMissing(data)
```

### Data Distribution

#### Continuous Features

###### Histogram
The following graph shows the density distribution of all continuous features.
```{r histogram_continuous}
HistogramContinuous(data)
```

###### Correlation Heatmap
```{r correlation_continuous}
CorrelationContinuous(data, use="na.or.complete")
```

#### Discrete Features
###### Bar Charts
The following graph shows the frequency distribution of all discrete features.
```{r bar_discrete}
BarDiscrete(data)
```

###### Correlation Heatmap
```{r correlation_discrete}
CorrelationDiscrete(data, use="pairwise.complete.obs")
```

